---
title: "ESM 270 Week 8: Analytic Hierarchy Process (AHP)"
subtitle: "Eliciting and mapping preferences for conservation priorities" 
author: "Cameryn Brock"
date: "10/06/2020"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

```{r}
library(here)
library(janitor)
library(sf)
library(ggmap)
library(tidyverse)
library(ahp)
library(yaml)
library(data.tree)
```

# Introduction

## AHP

Analytic Hierarchy Process (AHP) is a common approach to eliciting expert or stakeholder input on the importance of difference conservation factors.

## Data

We will be using a case study (and the data they prepared) from a past group project on the Morro Bay National Estuary Program.

# AHP

```{r}
c(
"Version: 2.0",
"Goal:",
"  name: element",
"  decision-makers:",
    #optional node, needed only if not all decision-makers have equal voting power
"    - User 1: 0.2",
"    - User 2: 0.2",
"    - User 3: 0.2",
"    - User 4: 0.2",
"    - User 5: 0.2",
"  preferences:",
    #fill in x's for user preferences
"    User 1:",
"      pairwise:",
"      - [Threatened/Endangered, Charismatic, 9]",
"      - [Threatened/Endangered, Migratory, 5]",
"      - [Threatened/Endangered, Sig. Economic Value, 3]",
"      - [Charismatic, Migratory, 1/5]",
"      - [Charismatic, Sig. Economic Value, 1/3]",
"      - [Migratory, Sig. Economic Value, 1]",
"    User 2:",
"      pairwise:",
"      - [Threatened/Endangered, Charismatic, 6]",
"      - [Threatened/Endangered, Migratory, 5]",
"      - [Threatened/Endangered, Sig. Economic Value, 3]",
"      - [Charismatic, Migratory, 5]",
"      - [Charismatic, Sig. Economic Value, 1/7]",
"      - [Migratory, Sig. Economic Value, 1/7]",
"    User 3:",
"      pairwise:",
"      - [Threatened/Endangered, Charismatic, 1/8]",
"      - [Threatened/Endangered, Migratory, 7]",
"      - [Threatened/Endangered, Sig. Economic Value, 1/9]",
"      - [Charismatic, Migratory, 7]",
"      - [Charismatic, Sig. Economic Value, 1/5]",
"      - [Migratory, Sig. Economic Value, 1/8]",
"    User 4:",
"      pairwise:",
"      - [Threatened/Endangered, Charismatic, 9]",
"      - [Threatened/Endangered, Migratory, 3]",
"      - [Threatened/Endangered, Sig. Economic Value, 3]",
"      - [Charismatic, Migratory, 1/7]",
"      - [Charismatic, Sig. Economic Value, 1/5]",
"      - [Migratory, Sig. Economic Value, 5]",
"    User 5:",
"      pairwise:",
"      - [Threatened/Endangered, Charismatic, 7]",
"      - [Threatened/Endangered, Migratory, 3]",
"      - [Threatened/Endangered, Sig. Economic Value, 4]",
"      - [Charismatic, Migratory, 1/7]",
"      - [Charismatic, Sig. Economic Value, 1/4]",
"      - [Migratory, Sig. Economic Value, 3]",
"  children:",
"    Threatened/Endangered:",
"    Charismatic:",
"    Sig. Economic Value:",
"    Migratory:") %>% 
  write_lines("ahp.txt")

ahp <- Load("ahp.txt") 

Calculate(ahp)

ahp_weights <- print(ahp, priority = function(x) x$parent$priority["Total", x$name]) %>% 
  rename(element = levelName) %>% 
  drop_na() %>% 
  mutate(element = str_sub(element, 5))
```

## Calculate weighted sums with AHP output

Prepare a file with your target species and their status (1 or 0) for the focal elements: 

* Threatened/Endangered
* Charismatic
* Migratory
* Sig. Economic Value

We are specifically interested in the cumulative weighted sums for each species, so we will combine your file with the weights calculated above.  

```{r}
spec_list <- read_csv(here("data", "spec-list.csv")) %>% 
  clean_names()

spec_list

weights_long <- ahp_weights %>% 
  pivot_wider(names_from = element,
              values_from = priority) %>% 
  clean_names()

spec_weighted <- spec_list %>% 
  mutate(threatened_endangered = threatened_endangered * weights_long$threatened_endangered,
         charismatic = charismatic * weights_long$charismatic,
         migratory = migratory * weights_long$migratory,
         sig_economic_value = sig_economic_value * weights_long$sig_economic_value)

# calculate weighted sums

spec_sums <- spec_weighted %>% 
  group_by(common_name, latin_name) %>% 
  summarize(weighted_sum = sum(threatened_endangered, charismatic, 
                               migratory, sig_economic_value))
```

# Visualize Results

## Read in Morro Bay data

```{r}
# species and pu data
# all csv files are unaltered from the xlsx files in the R drive. Just saved as csvs.

spec <- read_csv(here("data","morro-bay-spec.csv")) %>% 
  head(140) %>% 
  rename(latin_name = name)

puvsp <- read_csv(here("data","morro-bay-puvspr.csv")) %>% 
  select(1:3) %>% 
  head(11849)

# polygons

parcels <- read_sf(dsn = here("data"), layer = "MorroBay_parcels") %>% 
  clean_names()
```

## Combine weight values with planning units and sum based on pu 

```{r}
spec_weight <- inner_join(spec_sums, spec, by = "latin_name") %>% 
  rename(species = id) %>% 
  select(species, common_name, latin_name, weighted_sum)

puvsp_weight <- left_join(puvsp, spec_weight, by = "species") 

pu_weight <- puvsp_weight %>% 
  group_by(pu) %>% 
  summarize(cum_sum = sum(weighted_sum,
                          na.rm = TRUE)) %>% 
  rename(id = pu)
```

## Bind cumulative sums to parcel shapes

```{r}
shp_weight <- inner_join(parcels, pu_weight, by = "id") %>% 
  select(id, cum_sum) # geometry is sticky and still in df
```

## Plot!

```{r}
ggplot(data = shp_weight) + 
  geom_sf(aes(fill = cum_sum),
          color = "white",
          size = 0.05) +
  scale_fill_continuous(low = "slategray2",
                        high = "navy") +
  labs(fill = "Cumulative \nWeighted Sum") +
  theme_minimal()
```

## Visualize with basemap from `ggmap`

```{r}
# get basemap with ggmaps

morrobay <- get_map(location = c(lon = -120.7665, lat = 35.335),
                    zoom = 12,
                    maptype = "terrain-background", 
                    source = "google")

ggmap(morrobay) +
  geom_sf(data = shp_weight,
          aes(fill = cum_sum),
          color = "white",
          size = 0.1,
          alpha = 0.85,
          inherit.aes = FALSE) + 
  coord_sf(crs = st_crs(4326)) + 
  scale_fill_gradient(low = "slategray2",
                      high = "navy") +
  labs(fill = "Cumulative \nWeighted Sum",
       x = NULL,
       y = NULL) +
  theme_minimal() 

ggsave("week-8-ahp.png")
```

